A Triple Origin for Twin Blue Stragglers in Compact Binaries

### 2018-12-10 BSS7782.tex
Update BSS7782.tex

We propose a formation mechanism for twin blue stragglers in compact
binaries that involves mass transfer from an evolved outer tertiary
companion on to the inner binary via a circumbinary disk.  We apply
our hypothetical scenario to the observed double blue straggler system
Binary 7782 in the old open cluster NGC 188, and show that its
observed properties are naturally reproduced within the context our
proposed model.  Within the context of the hypothetical formation
mechanism proposed here, the presented work predicts the following
properties for the post-mass transfer double BS tertiary: (1) For the
outer tertiary orbit, the initial orbital period should lie between
220 days $\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days, assuming
initial masses for the inner binary components of m$_{\rm 1} =$ 1.1
M$_{\odot}$ and m$_{\rm 2} =$ 0.9 M$_{\odot}$ and an outer tertiary
mass of m$_{\rm 3} =$ 1.4 M$_{\odot}$. (2) Larger final WD masses, and
hence core masses for the donor at the time of mass transfer, should
correspond to larger final outer orbital periods for the tertiary. (3)
For the inner binary, the rotational axes of the BSs should be aligned
with each other and the orbital plane of the outer tertiary WD. (4)
The BSs in the inner binary should have roughly equal masses,
independent of their initial masses.  This predicts that the initially
lower mass MS star should accrete the most, and should hence be
polluted more effectively by the accreted material.  This could be
observable in the surface layers of a radiative star (i.e., He, C and
O if the donor is an RGB star, and/or s-process elements if the donor
is an AGB star). (5) Twin BSs in compact binaries formed from the
proposed mechanism should be more frequent in younger clusters with
ages $\lesssim$ 4-6 Gyr, since the donor will have radiative envelope.

@@

Even more curious, there exists in the old open cluster NGC 188 a
double BS binary, called Binary 7782.  Specifically, \citet{mathieu09}
observed a compact and mildly eccentric (i.e., e $\sim$ 0.1) binary
star system with an orbital period of $\sim$ 10 days hosting two blue
stragglers.  During a given binary-binary interaction, the probability
that not one but two direct (ms, ms) collisions occur is less than
$10^{-2}$ \citep{leonard89,leigh11,leigh12}.  Typically, such binaries
have long orbital periods. Dynamically, it difficult to form a
short-period binary composed of two collision products during a
collision interaction in a star cluster \citep{2011Sci...334.1380F}.
So, how did Binary 7782 form?  MENTION BIMODAL P-E DISTRIBUTION, AND
HOW THIS COULD BE SUGGESTIVE OF A TRIPLE ORIGIN FOR SOME OF THE SUBSET
OF OBSERVED BSs.

@@

Consider a hierarchical triple system with component masses m$_{\rm
  1}$ and m$_{\rm 2}$ for the inner binary, and mass m$_{\rm 3}$ for
the outer tertiary companion.  The inner and outer binary orbital
semi-major axes are denoted a$_{\rm in}$ and a$_{\rm out}$,
respectively.  We assume circular orbits for both the inner and outer
orbits, and co-planar triples only since this is what has been
observed for low-mass tertiaries \cite{2010yCat..73890925T} \simon{I
  think that {\em moe18} and {\em tobin18} are not the right
  references}.  This initial configuration for our assumed formation
scenario for Binary 7782, described below, is depicted in
Figure~\ref{fig:fig1}.

@@

We consider a scenario where the outer star is filling its Roche lobe
and is transfers mass to the inner binary.  The mass transfer stream
gathers at the circularization radius a$_{\rm c}$, and forms a disk
around the inner binary.  Using conservation of angular momentum, we
equate the specific angular momentum of the accreted mass at the inner
Lagrangian point of the (outer) donor star to the final specific
angular momentum of the accretion stream at the circularization radius
about the inner binary, this results in
\simon{replaced lower case $r$ with upper case.}
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - R_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer
tertiary companion, a$_{\rm c}$ is the semi-major axis of the orbit
about the inner binary corresponding to the circularization radius and
v$_{\rm orb,c}$ is the orbital velocity at a$_{\rm c}$.  The distance
from the center of mass corresponding to the outer tertiary companion
defined by the Roche lobe is given by \citep{eggleton83}.
\simon{Do we really need this eq?}
\begin{equation}
\label{eqn:roche}
R_{\rm L} = \frac{0.49q^{2/3}}{0.6q^{2/3} + {\rm ln}(1 + q^{1/3})}a_{\rm out},
\end{equation}
where the mass ratio q is defined as q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$
m$_{\rm 2}$).  Combining Equation~\ref{eqn:roche} with
Equation~\ref{eqn:specangmom1}, we can solve for the circularization
radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we
require that a$_{\rm in} <$ a$_{\rm c}$.

Figure~\ref{fig:fig2} shows the parameter space in the P$_{\rm
  out}$-P$_{\rm in}$-plane for Binary 7782.  We assume initial
component masses of m$_{\rm 1} =$ 1.1 M$_{\rm \odot}$ and m$_{\rm 2}
=$ 0.9 M$_{\rm \odot}$ for the inner binary components, and m$_{\rm 3}
=$ 1.4 M$_{\rm \odot}$ for the outer tertiary.  We compare the
circularization radius to the semi-major axis of the inner binary, for
which we require a$_{\rm c} >$ a$_{\rm in}$, \del{after folding in all
  constraints from the requirements for dynamical stability}\simon{can
  we not simply state the constraints?}, an outer tertiary that is
Roche lobe-filling, and a dynamically hard outer tertiary orbit.  Note
that the range of plotted orbital periods P$_{\rm in}$ corresponding
to a contact state for the inner binary, assuming \simon{This looks
  weird:} R$_{\rm 1} =$ R$_{\rm 2} =$ 1 R$_{\odot}$, lies outside the
range of plotted values for P$_{\rm in}$, since it does not contribute
significantly to constraining the outer orbital properties of a
hypothesized outer WD tertiary.  The thick horizontal solid red line
shows the allowed range of outer semi-major axes, after folding in all
of the aforementioned criteria.  As is clear, this makes a relatively
narrow prediction for the allowed ranges of outer tertiary orbits,
namely 2.2 $\times$ 10$^{2}$ days $\le$ P$_{\rm out}$ $\le$ 1.1
$\times$ 10$^3$ days, for our assumed final donor mass.

\begin{figure}[ht!]
\plotone{fig2.eps}
\caption{Parameter space in the P$_{\rm out}$-P$_{\rm in}$-plane
  allowed for the hypothetical outer tertiary orbit of Binary 7782.
  The solid diagonal black lines show the period corresponding to the
  circularization radius a$_{\rm c}$ for the mass transfer stream
  coming from the outer star, for different values of the mass ratio,
  namely q $=$ 0.1, 0.7, 1, 2 and 10 \simon{before or after mass
    transfer?}.  \simon{what is the purpose of extending q to 0.1 and
    10?} We assume initial component masses of m$_{\rm 1}$ = 1.1
  M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner
  binary components, and m$_{\rm 3}$ is computed for the outer
  tertiary according to our assumed mass ratio (with our fiducial case
  corresponding to q $=$ 0.7.  We assume completely conservative mass
  transfer for this exercise, and a final mass for the outer tertiary
  of 0.6 M$_{\rm \odot}$ once it has become a WD.  The dashed diagonal
  black line shows a rough criterion for dynamical stability in the
  triple, approximately following \citet{mardling99} (i.e., a$_{\rm
    in} <$ 0.1a$_{\rm out}$ is required for long-term dynamical
  stability in equal-mass co-planar triples).  The vertical solid red
  lines show the outer orbital periods corresponding to the hard-soft
  boundary assuming central velocity dispersions of $\sigma =$ 1, 5
  and 10 km s$^{-1}$.  The vertical dashed black lines show the
  maximum outer orbital period P$_{\rm out}$ for which the outer
  tertiary companion is Roche lobe-filling, assuming a stellar radius
  of R$_{\rm 3} =$ 200 R$_{\odot}$ (CHANGE, POSSIBLY USING MESA
  CALCULATIONS?!).  The horizontal dashed red line shows the observed
  orbital period for Binary 7782, using its observed orbital period
  and our assumed final inner companion masses (i.e., m$_{\rm 1} =$
  m$_{\rm 2} =$ 1.4 M$_{\odot}$).  Finally, the thick solid horizontal
  red line shows the parameter space for P$_{\rm out}$ allowed after
  considering all of the aforementioned criteria.  \simon{I think that
    the figure needs work. I like putting the parameter space in terms
    of orbital periods, but the q=0.1 and q=10 constraints seems
    not very practical. The red labels are not readable, and I am not
    really sure how to read the figure. What about the alternative in
    the next figure, which shows tertiary mass vs. orbital period?}
\label{fig:fig2}}
\end{figure}

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_minimumstablesize.pdf}
\caption{Orbital separation as a function of the mass of the
  Roche-lobe filling outer star. The red curves gives the size at the
  first giant branch (thick curve) at at the AGB (thin curve). The
  blue curve the orbital separation at the moment this giant fills its
  Roche-lobe assuming a circular orbit and a total inner-binary mass
  of 2.0\,\MSun. The green curves give, under these conditions, the
  maximum orbital separation for a stable inner binary.
\label{fig:tertiarymass_vs_size}}
\end{figure}

>>>>

We propose a formation mechanism for twin blue stragglers in compact
binaries that involves mass transfer from an evolved outer tertiary
companion on to the inner binary via a circumbinary disk.  We apply
our hypothetical scenario to the observed double blue straggler system
Binary 7782 in the old open cluster NGC 188, and show that its
observed properties are naturally reproduced within the context our
proposed model.  Within the context of the hypothetical formation
mechanism proposed here, the presented work predicts the following
properties for the post-mass transfer double BS tertiary: (1) For the
outer tertiary orbit, the initial orbital period should lie between
220 days $\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days, assuming
initial masses for the inner binary components of m$_{\rm 1} =$ 1.1
M$_{\odot}$ and m$_{\rm 2} =$ 0.9 M$_{\odot}$ and an outer tertiary
mass of m$_{\rm 3} =$ 1.4 M$_{\odot}$. (2) Larger final WD masses, and
hence core masses for the donor at the time of mass transfer, should
correspond to larger final outer orbital periods for the tertiary. (3)
For the inner binary, the rotational axes of the BSs should be aligned
with each other and the orbital plane of the outer tertiary WD. (4)
The BSs in the inner binary should have roughly equal masses,
independent of their initial masses.  This predicts that the initially
lower mass MS star should accrete the most, and should hence be
polluted more effectively by the accreted material.  This could be
observable in the surface layers of a radiative star (i.e., He, C and
O if the donor is an RGB star, and/or s-process elements if the donor
is an AGB star). (5) Twin BSs in compact binaries formed from the
proposed mechanism should be more frequent in younger clusters with
ages $\lesssim$ 4-6 Gyr, since the donor will have a radiative envelope.

@@

Even more curious, there exists in the old open cluster NGC 188 a
double BS binary, called Binary 7782.  More generally, the BS population in NGC 188 has a bi-modal period-eccentricity distribution.  As discussed in \citet{leigh11}, this could be hinting at a triple origin for at least some subset of the total BS population.  As for Binary 7782, \citet{mathieu09}
observed a compact and mildly eccentric (i.e., e $\sim$ 0.1) binary
star system with an orbital period of $\sim$ 10 days hosting two blue
stragglers.  During a given binary-binary interaction, the probability
that not one but two direct (ms, ms) collisions occur is less than
$10^{-2}$ \citep{leonard89,leigh11,leigh12}.  Typically, such binaries
have long orbital periods. Dynamically, it is difficult to form a
short-period binary composed of two collision products during a
collision interaction in a star cluster \citep{2011Sci...334.1380F}.
So, how did Binary 7782 form?

@@

Consider a hierarchical triple system with component masses m$_{\rm
  1}$ and m$_{\rm 2}$ for the inner binary, and mass m$_{\rm 3}$ for
the outer tertiary companion.  The inner and outer binary orbital
semi-major axes are denoted a$_{\rm in}$ and a$_{\rm out}$,
respectively.  We assume circular orbits for both the inner and outer
orbits, and co-planar triples only since this is what has been
observed for low-mass tertiaries \cite{2010yCat..73890925T} \simon{I
  think that {\em moe18} and {\em tobin18} are not the right
  references...NL:  Agreed.  Do you have others to recommend?  These were what came to mind, but 
  I can do some digging to try to find the more appropriate refs.}.  This initial configuration for our assumed formation
scenario for Binary 7782, described below, is depicted in
Figure~\ref{fig:fig1}.

@@

We consider a scenario where the outer star is filling its Roche lobe
and transfers mass to the inner binary.  The mass transfer stream
gathers at the circularization radius a$_{\rm c}$, and forms a disk
around the inner binary.  Using conservation of angular momentum, we
equate the specific angular momentum of the accreted mass at the inner
Lagrangian point of the (outer) donor star to the final specific
angular momentum of the accretion stream at the circularization radius
about the inner binary, this results in
\simon{replaced lower case $r$ with upper case.}
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - R_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer
tertiary companion, a$_{\rm c}$ is the semi-major axis of the orbit
about the inner binary corresponding to the circularization radius and
v$_{\rm orb,c}$ is the orbital velocity at a$_{\rm c}$.  The distance
from the center of mass corresponding to the outer tertiary companion
defined by the Roche lobe is given by Equation 2 in \citep{eggleton83}.
\simon{Do we really need this eq? - NL: Good call, no.  We can definitely remove this to save space, given that this is a Letter.  I have done so.}
%\begin{equation}
%\label{eqn:roche}
%R_{\rm L} = \frac{0.49q^{2/3}}{0.6q^{2/3} + {\rm ln}(1 + q^{1/3})}a_{\rm out},
%\end{equation}
%where the mass ratio q is defined as q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$
%m$_{\rm 2}$).  
Combining Equation 2 in \citet{eggleton83} (with mass ratio q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$m$_{\rm 2}$)) with
Equation~\ref{eqn:specangmom1}, we can solve for the circularization
radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we
require that a$_{\rm in} <$ a$_{\rm c}$.

Figure~\ref{fig:fig2} shows the parameter space in the P$_{\rm
  out}$-P$_{\rm in}$-plane for Binary 7782.  We assume initial
component masses of m$_{\rm 1} =$ 1.1 M$_{\rm \odot}$ and m$_{\rm 2}
=$ 0.9 M$_{\rm \odot}$ for the inner binary components, and m$_{\rm 3}
=$ 1.4 M$_{\rm \odot}$ for the outer tertiary.  We compare the
circularization radius to the semi-major axis of the inner binary, for
which we require a$_{\rm c} >$ a$_{\rm in}$, \del{after folding in all
  constraints from the requirements for dynamical stability (listed in the caption of Figure~\ref{fig:fig2})}\simon{can
  we not simply state the constraints? - We do, but it is in the caption to save space.  It is all there anyways, so can be brought in to the main text if necessary.}, an outer tertiary that is
Roche lobe-filling, and a dynamically hard outer tertiary orbit.  Note
that the range of plotted orbital periods P$_{\rm in}$ corresponding
to a contact state for the inner binary lies outside the
range of plotted values for P$_{\rm in}$ (for components with radii of roughly 1 R$_{\odot}$), since it does not contribute
significantly to constraining the outer orbital properties of a
hypothesized outer WD tertiary.  The thick horizontal solid red line
shows the allowed range of outer semi-major axes, after folding in all
of the aforementioned criteria.  As is clear, this makes a relatively
narrow prediction for the allowed ranges of outer tertiary orbits,
namely 2.2 $\times$ 10$^{2}$ days $\le$ P$_{\rm out}$ $\le$ 1.1
$\times$ 10$^3$ days, for our assumed final donor mass.

\begin{figure}[ht!]
\plotone{fig2.eps}
\caption{Parameter space in the P$_{\rm out}$-P$_{\rm in}$-plane
  allowed for the hypothetical outer tertiary orbit of Binary 7782.
  The solid diagonal black lines show the period corresponding to the
  circularization radius a$_{\rm c}$ for the mass transfer stream
  coming from the outer star, for different values of the mass ratio,
  namely q $=$ 0.1, 0.7, 1, 2 and 10 \simon{before or after mass
    transfer?}.  \simon{what is the purpose of extending q to 0.1 and
    10? - NL: Fair enough.  It was arbitrary, admittedly, but supposed to take in to account the fact that the mass ratio changes during MT.  I suppose I thought it illustrated that nothing much changes upon changing q, hence the conclusions are robust over the entire extent of the MT process, during which the mass ratio of course changes.} We assume initial component masses of m$_{\rm 1}$ = 1.1
  M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner
  binary components, and m$_{\rm 3}$ is computed for the outer
  tertiary according to our assumed mass ratio (with our fiducial case
  corresponding to q $=$ 0.7).  We assume completely conservative mass
  transfer for this exercise, and a final mass for the outer tertiary
  of 0.6 M$_{\rm \odot}$ once it has become a WD.  The dashed diagonal
  black line shows a rough criterion for dynamical stability in the
  triple, approximately following \citet{mardling99} (i.e., a$_{\rm
    in} <$ 0.1a$_{\rm out}$ is required for long-term dynamical
  stability in equal-mass co-planar triples).  The vertical solid red
  lines show the outer orbital periods corresponding to the hard-soft
  boundary assuming central velocity dispersions of $\sigma =$ 1, 5
  and 10 km s$^{-1}$.  The vertical dashed black lines show the
  maximum outer orbital period P$_{\rm out}$ for which the outer
  tertiary companion is Roche lobe-filling, assuming a stellar radius
  of R$_{\rm 3} =$ 200 R$_{\odot}$ (CHANGE, POSSIBLY USING MESA
  CALCULATIONS?!).  The horizontal dashed red line shows the observed
  orbital period for Binary 7782, using its observed orbital period
  and our assumed final inner companion masses (i.e., m$_{\rm 1} =$
  m$_{\rm 2} =$ 1.4 M$_{\odot}$).  Finally, the thick solid horizontal
  red line shows the parameter space for P$_{\rm out}$ allowed after
  considering all of the aforementioned criteria.  \simon{I think that
    the figure needs work. I like putting the parameter space in terms
    of orbital periods, but the q=0.1 and q=10 constraints seems
    not very practical. The red labels are not readable, and I am not
    really sure how to read the figure. What about the alternative in
    the next figure, which shows tertiary mass vs. orbital period?}
\label{fig:fig2}}
\end{figure}

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_minimumstablesize.pdf}
\caption{Orbital separation as a function of the mass of the
  Roche-lobe filling outer star. The red curves gives the size at the
  first giant branch (thick curve) at the AGB (thin curve). The
  blue curve gives the orbital separation at the moment this giant fills its
  Roche-lobe assuming a circular orbit and a total inner binary mass
  of 2.0\,\MSun. The green curves give, under these conditions, the
  maximum orbital separation for a stable inner binary.
\label{fig:tertiarymass_vs_size}}
\end{figure}
